Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-04 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Draft for final feedback

\begin{abstract}
Iterative imputation is a popular tool to accommodate missing data. While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence. There is no consensus on how to evaluate the convergence properties of the method. This paper provides insight into identifying non-convergence of iterative impuation algorithms.
\end{abstract}

@@

Currently, the recommended practice for evaulating the convergence of iterative imputation algorithms is through visual inspection. That is, the state space of the algorithm is monitored by plotting a scalar summary \(\theta\) against the iteration number.
\textbf{{[}Segway to WHAT TO MONITOR? Also, this paragraph is more or less the same as the third paragraph in the second section. Merge or remove one of the two?{]}} Typically, the \(\theta\)s under evaluation are chain means and variances (\textbf{define!}). These \(\theta\)s may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of \(y_{imp}\). A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient). Implementing this, however, might be somewhat advanced for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest. Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently. Therefore, this \(\theta\) is not sufficient either. Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}. Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance. This would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented.
\textbf{{[}Segway to HOW TO MONITOR?{]}} Monitoring convergence visually is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward. Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed. First, we define which non-convergence diagnostics will be considered, and evaluate how these diagnostics could be appropriate for iterative imputation applications. In the second part of this paper, we will investigate the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in \texttt{R} \citep{R}. The aim of the simulation is to determine whether unbiased, confidence valid inferences may be obtained under different levels of non-convergence (\textbf{and at which point convergence may safely be assumed}). Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence. The results of this study are guidelines for identifying non-convergence in iterative imptation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets. \textbf{{[}add more societal relevance: what is the consequence of non-convergence? e.g., biased estimates, invalid inference, etc.{]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on \(k\) variables for all \(n\) units in a sample. The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing. The number of units \(i\) in dataset \(y\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{mis}\). The collection of observed data values in \(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample \(M \times T\) plausible values, where \(M\) is the number of imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm. The collection of samples between the initial value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'. \textbf{\(\theta\)s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain). }

@@

\textbf{{[}Move to INTRO???{]}} To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call \(j\). We see two traceplots of the chain means for \(j\)---i.e., the average of the imputed values for \(j\) in each imputation \(y_{imp,m}\), plotted against the iteration number. Each line portrays one imputation. The plot on the lefthand side of the figure shows typical convergence of an iterative imputation algorithm. The righthand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. \textbf{{[}Leave this in if it's after mixing and stationarity are defined, otherwise remove?{]} In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence plot, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per \(m\)) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. That means that non-convergence has a impact on the mean of \(j\) in \(y_{imp, m}\). This difference (presumably) translates to bias in the \(M\) completed data estimates \(\hat{\mu}_{j,m}\), and consequently also to the pooled estimate \(\bar{\mu}_j\). Non-convergence thus leads to biased estimates. This shows the importance of reaching converged states in iterative imputation algorithms.

@@

\textbf{{[}Move to INTRO???{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteraton number is plotted against a certain \(\theta\). \(\theta\)s are scalar summaries of the state space of the algorithm at a specific iteration. The recommended scalar summaries to evaluate are chain means and chain variances. Additionally, researchers may ``monitor some statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}. As \citet{buur18} describes, researchers can specify their quantity of scientific interest \(Q\) (e.g., a regression coefficient) as scalar summary \(\theta\). Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers. And moreover, it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. Therefore, we propose a \(\theta\) that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest. We propose \(\lambda_1\) as such a scalar summary. We define \(\lambda_1\) as the first eigenvalue of the variance-covariance matrix of the completed data. Let \(\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j\) be the eigenvalues of \(\Sigma\) in each of the \(M\) completed datasets \({y_{obs}, y_{imp}}\). \(\lambda_1\) is measure that summarises the covariances in the completed datasets. \textbf{The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by \citet{mack03}.}

@@

Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for converegnce: mixing or stationarity. Non-stationarity within chains may be diagnosed with e.g., autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's \citeyearpar{raft91} procedure to determine the effect of trending on the precision of estimates. A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor \(\widehat{R}\) \citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed adaptation, \(\widehat{R}\) might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated \citep{veht19}. Therefore, we use \(\widehat{R}\) and \(AC\) to evaluate mixing and stationarity separately, as recommended by e.g., \citet{cowl96}.

@@

An updated version of \(\widehat{R}\) has been proposed by Vehtari et al.~\citeyearpar[p.~5]{veht19}. \textbf{This version may be suitable for iterative imputation.} Let \(M\) be the total number of chains, \(T\) the number of iterations per chain, and \(\theta\) the scalar summary of interest (e.g., chain means). For each chain (\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).

@@

We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\) per chain.

@@

From the between- and within-chain variances we compute a weighted average, \(\widehat{\operatorname{var}}^{+}\), which over-estimates the total variance of \(\theta\) \textbf{remove or explain why, or leave in}. \(\widehat{R}\) is then obtained as a ratio between the over-estimated total variance and the within-chain variance:

@@

We can interpret \(\widehat{R}\) as potential scale reduction factor since it indicates by how much the variance of \(\theta\) could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}. This interpretation assumes that chains are `over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\). Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other. When all chains sample independent of their initial values, the mixing component of convergence is satisfied, and \(\widehat{R}\)-values will be close to one. High \(\widehat{R}\)-values thus indicate non-convergence. The conventionally acceptable threshold for convergence was \(\widehat{R} < 1.2\) \citep{gelm92}. More recently, \citet{veht19} proposed a more stringent threshold of \(\widehat{R} < 1.01\).

@@

Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}=1\) and \(AC=0\) quantitatively through simulation, we apply them to the example of pathological non-convergence by \citet{buur18} that we reproduced above. We want to be able to draw the same conclusions as we did from visually inspecting the two algorithms. But \(\widehat{R}=1\) and \(AC=0\) should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}. From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around \(t=2\)), but intermigled nicely overall. Additionally, there was very little trending. The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from \(t=7\)). Moreover, there was a lot of trending as well initially (up-to \(t=6\)), after which the chains reached a somewhat more stationary state. To assess whether \(\widehat{R}\) and \(AC\) may be appropriate non-convergence identifiers for iterative imputation algorithms, they should identify the same trends as the visual inspection. But at least, the following condition must hold: the methods should indicate worse performance for the mis-specified model with pathological non-convergence than the typically converged algorithm (i.e., higher \(\widehat{R}\)- and \(AC\)-values).

In Figure \ref{fig:diagnostics}A, the chain means from Figure \ref{fig:non-conv} are plotted again---now together. Panel B shows two versions of calculating \(AC\)s applied to the \(\theta\)s of panel A: the default calculation with \texttt{R} function \texttt{stats::acf()} \citep{R}, and manual calculation as the correlation between \(\theta\) in iteration \(t\) and \(\theta\) in iteration \(t+1\). Panel C shows the traditional computation of \(\widehat{R}\), and panel D shows \(\widehat{R}\) as computed by implementing \citet{veht19} 's recommendations.

@@

\textbf{Describe panel C here and why it is not optimal{]}} The \(\widehat{R}\)-values in panel D do meet the requirements specified above. \(\widehat{R}\) as computed conform \citet{veht19} indeed indicate less signs of non-convergence as the number of iterations goes up.

@@

Non-convergence will be induced by: 1) increasing the missingness proportion \(p_{mis}\) in dataset \(y\) incrementally, and 2) terminating the imputation algorithm at a varying number of iterations \(T\). The first set of simulation conditions---the missingness proportions---is chosen to reflect the difficulty of the missingness problem. The underlying assumption is that low missingness proportions lead to quick algorithmic convergence, since there is a lot of information in the observed data. Higher missingness proportions should yield slower convergence. Unless, however, the fraction of missing information is so high that the random component in the imputation algorithm outweighs the information in the observed data. Then, convergence to a stable but highly variable state may be reached instanteneously. We assume that the incremental missingness proportions in our study will result in a corresponding increase in signs of non-convergence.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We expect that simulation conditions with a high missingness proportion \(p_{mis}\) and a low number of iterations \(T\) will result in biased, invalid estimates of the quantities of scientific interest, \(Q\)s.
\item
  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify signs of non-convergence in those simulation conditions where the \(\bar{Q}\)s are \emph{not} unbiased and confidence valid estimates of the \(Q\)s.
\item
  We hypothesize that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\), \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too stringent for iterative imputation applications. In an empirical study, where \(\widehat{R}\) was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet, scientific estimates were insensitive to continued iteration from \(t=5\) onwards \citep{lace07}. We therefore suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the bias the estimates. In contrast to this, it may occur that signs of non-convergence are under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values. In practice, we do not expect this to cause problems for identifying non-convergence with \(\widehat{R}\).
\item
  We expect that high \(AC\) values are implausible in iterative imputation algorithms with typical convergence. That is, after only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
\item
  \textbf{Something about thetas}
\end{enumerate}

@@

In this study, sampling variance is not of interest. Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions \citep{vink14}. The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem \textbf{(from an unspecified social scientific field of study)}. Let the predictor space be defined as
\[
\begin{aligned}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{aligned}
\]
A finite population of \(N=1000\) is simulated using the \texttt{mvtnorm} package \citep{mvtnorm}. Subsequently, a fourth variable is constructed as outcome \(Y\). For each unit \(i = 1, 2,..., N\), let
\[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\]
where \(\epsilon \sim \mathcal{N}(0, 100)\). From the complete set, we obtain the true values of several quantities of scientific interest, \(Q\)s. \textbf{{[}MOVE THIS NEXT PART TO TOP OF SIMULATION SECTION?{]}} \(Q\)s are the targets that will be estimated in each simulation repetition, after first \emph{amputing} the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the \(Q\)s on the \(M\) completed datasets, and finally pooling the \(M\) estimates \(\hat{Q}\) to obtain a single \(\bar{Q}\) for each \(Q\). \textbf{We try to predict the effect of non-converegence (i.e., the difference between \(Q\) and \(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and \(AC\), by appliying these methods to several scalar summaries of the state space of the algorithm, \(\theta\)s.}

@@

\textbf{{[}Add to this section: 1) emphasize that the iterations in the plots are averages across repetitions, not within one run of MICE; 2) sample effects due to single complete dataset?? 3) Add more info about figure legends and axes 4) the ac and rhat are independent of missingness, not the thetas.{]}}
Primarily, we want to know what the effect is of inducing non-convergence on the estimated \(Q\)s. Specifically, we are interested in the bias of \(\bar{Q}\) for all \(Q\)s and the confidence validity of \(\bar{Q}\) for \(Q=\beta\). Therefore, we plot these performance measures against the simulation conditions (the number of iterations \(T\) and the missingness proportion \(p_{mis}\)). We look at averages across simulation repetitions.

>>>>

\begin{abstract}
Iterative imputation is a popular tool to accommodate missing data. While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence. There is no consensus on how to evaluate the convergence properties of the method. This paper provides insight into identifying non-convergence of iterative imputation algorithms.
\end{abstract}

@@

Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection. That is, the state space of the algorithm is monitored by plotting a scalar summary \(\theta\) against the iteration number.
\textbf{{[}Segway to WHAT TO MONITOR? Also, this paragraph is more or less the same as the third paragraph in the second section. Merge or remove one of the two?{]}} Typically, the \(\theta\)s under evaluation are chain means and variances (\textbf{define!}). These \(\theta\)s may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with univariate estimates, but the entire multivariate distribution of \(y_{imp}\). A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient). Implementing this, however, might be somewhat advanced for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest. Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently. Therefore, this \(\theta\) is not sufficient either. Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}. Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance. This would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented.
\textbf{{[}Segway to HOW TO MONITOR?{]}} Monitoring convergence visually is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward. Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed. First, we define which non-convergence diagnostics will be considered, and evaluate how these diagnostics could be appropriate for iterative imputation applications. In the second part of this paper, we will investigate the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in \texttt{R} \citep{R}. The aim of the simulation is to determine whether unbiased, confidence valid inferences may be obtained under different levels of non-convergence (\textbf{and at which point convergence may safely be assumed}). Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence (\textbf{and on which scalar summary of the state space of the algorithm these diagnostics should be applied}). The results of this study are guidelines for identifying non-convergence in iterative imputation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets. \textbf{{[}add more societal relevance: what is the consequence of non-convergence? see example of non-convergence in Figure 2 (e.g., biased estimates, invalid inference, etc.){]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on \(k\) variables for all \(n\) units in a sample. The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing. The number of units \(i\) in dataset \(y\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{mis}\). The collection of observed data values in \(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample \(M \times T\) plausible values, where \(M\) is the number of imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm. The collection of samples between the initial value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'. \textbf{{[}If not defined elsewhere, add: \(\theta\)s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).{]} }

@@

\textbf{{[}Move to INTRO???{]}} To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call \(j\). We see two traceplots of the chain means for \(j\)---i.e., the average of the imputed values for \(j\) in each imputation \(y_{imp,m}\), plotted against the iteration number. Each line portrays one imputation. The plot on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. \textbf{{[}Leave this in if it's after mixing and stationarity are defined, otherwise remove?{]} In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence plot, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per \(m\)) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. That means that non-convergence has a impact on the mean of \(j\) in \(y_{imp, m}\). This difference (presumably) translates to bias in the \(M\) completed data estimates \(\hat{\mu}_{j,m}\), and consequently also to the pooled estimate \(\bar{\mu}_j\). Non-convergence thus leads to biased estimates. This shows the importance of reaching converged states in iterative imputation algorithms.

@@

\textbf{{[}Move to INTRO???{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteration number is plotted against a certain \(\theta\). \(\theta\)s are scalar summaries of the state space of the algorithm at a specific iteration. The recommended scalar summaries to evaluate are chain means and chain variances. Additionally, researchers may ``monitor some statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}. As \citet{buur18} describes, researchers can specify their quantity of scientific interest \(Q\) (e.g., a regression coefficient) as scalar summary \(\theta\). Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers. And moreover, it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. Therefore, we propose a \(\theta\) that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest. We propose \(\lambda_1\) as such a scalar summary. We define \(\lambda_1\) as the first eigenvalue of the variance-covariance matrix of the completed data. Let \(\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j\) be the eigenvalues of \(\Sigma\) in each of the \(M\) completed datasets \({y_{obs}, y_{imp}}\). \(\lambda_1\) is measure that summarizes the covariances in the completed datasets. \textbf{The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by \citet{mack03}.}

@@

Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for convergence: mixing or stationarity. Non-stationarity within chains may be diagnosed with e.g., autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's \citeyearpar{raft91} procedure to determine the effect of trending on the precision of estimates. A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor \(\widehat{R}\) \citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed adaptation, \(\widehat{R}\) might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated \citep{veht19}. Therefore, we use \(\widehat{R}\) and \(AC\) to evaluate mixing and stationarity separately, as recommended by e.g., \citet{cowl96}.

@@

The potential scale reduction factor \(\widehat{R}\) was coined by \citet{gelm92}. An updated version of \(\widehat{R}\) has been proposed by Vehtari et al.~\citeyearpar[p.~5]{veht19}. \textbf{This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to. Namely, folded, rank-normalized, split-chain \(\theta\)s. The Vehtari et al.~version of \(\widehat{R}\) may be suitable for iterative imputation. We therefore use their definition of the diagnostic.} Let \(M\) be the total number of chains, \(T\) the number of iterations per chain, and \(\theta\) the scalar summary of interest (e.g., chain means). For each chain (\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).

@@

\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\) per chain.

@@

\noindent From the between- and within-chain variances we compute a weighted average, \(\widehat{\operatorname{var}}^{+}\), which approximates the total variance of \(\theta\). \(\widehat{R}\) is then obtained as a ratio between the total variance and the within-chain variance:

@@

We can interpret \(\widehat{R}\) as potential scale reduction factor since it indicates by how much the variance of \(\theta\) could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}. This interpretation assumes that chains are `over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\). Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other. When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied. The variance between chains is then equivalent to the variance within chains, and \(\widehat{R}\)-values will be close to one. High \(\widehat{R}\)-values thus indicate non-convergence. The conventionally acceptable threshold for convergence was \(\widehat{R} < 1.2\) \citep{gelm92}. More recently, \citet{veht19} proposed a more stringent threshold of \(\widehat{R} < 1.01\).

@@

Before we evaluate the performance of the non-convergence diagnostics \(\widehat{R}=1\) and \(AC=0\) quantitatively through simulation, we apply them to the example of pathological non-convergence by \citet{buur18} that we reproduced above. We want to be able to draw the same conclusions as we did from visually inspecting the two algorithms. But \(\widehat{R}=1\) and \(AC=0\) should at least be able to distinguish between the two algorithms plotted in Figure \ref{fig:non-conv}. From visual inspection, we know that the typically converged algorithm initially portrayed some signs of non-mixing (around \(t=2\)), but intermingled nicely overall. Additionally, there was very little trending. The algorithm with pathological non-convergence showed severe non-mixing, although this gradually improved (from \(t=7\)). Moreover, there was a lot of trending as well initially (up-to \(t=6\)), after which the chains reached a somewhat more stationary state. To assess whether \(\widehat{R}\) and \(AC\) may be appropriate non-convergence identifiers for iterative imputation algorithms, they should identify the same trends as the visual inspection. But at least, the following condition must hold: the methods should indicate worse performance for the mis-specified model with pathological non-convergence than the typically converged algorithm (i.e., higher \(\widehat{R}\)- and \(AC\)-values).

In Figure \ref{fig:diagnostics}A, the chain means from Figure \ref{fig:non-conv} are plotted again---now together. Panel B shows two versions of calculating \(AC\)s applied to the \(\theta\)s of panel A: the default calculation with \texttt{R} function \texttt{stats::acf()} \citep{R}, and manual calculation as the correlation between \(\theta\) in iteration \(t\) and \(\theta\) in iteration \(t+1\). Panel C shows the traditional computation of \(\widehat{R}\), and panel D shows \(\widehat{R}\) as computed by implementing \citet{veht19} 's recommendations. \textbf{{[}Add labels for the two types of Rhat to the figure!{]}}

@@

\textbf{Describe panel C here and why it is not optimal{]}} The \(\widehat{R}\)-values in panel D do meet the requirements specified above. \(\widehat{R}\) as computed conform \citet{veht19} does indeed indicate less signs of non-convergence as the number of iterations goes up. \textbf{\citep[explain the dip in Rhat values at t=2. Namely, because we can only use 2 of the 3 tricks by][ if the number of iterations is very low (t\textless4). That's why the \(\widehat{R}\)s are more similar to the traditional GR.]{veht19}}

@@

Non-convergence will be induced by: 1) increasing the missingness proportion \(p_{mis}\) in dataset \(y\) incrementally, and 2) terminating the imputation algorithm at a varying number of iterations \(T\). The first set of simulation conditions---the missingness proportions---is chosen to reflect the difficulty of the missingness problem. The underlying assumption is that low missingness proportions lead to quick algorithmic convergence, since there is a lot of information in the observed data. Higher missingness proportions should yield slower convergence. Unless, however, the fraction of missing information is so high that the random component in the imputation algorithm outweighs the information in the observed data. Then, convergence to a stable but highly variable state may be reached instantaneously. We assume that the incremental missingness proportions in our study will result in a corresponding increase in signs of non-convergence.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We expect that simulation conditions with a high missingness proportion \(p_{mis}\) and a low number of iterations \(T\) will result in biased, invalid estimates of the quantities of scientific interest, \(Q\)s.
\item
  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify signs of non-convergence in those simulation conditions where the \(\bar{Q}\)s are \emph{not} unbiased and confidence valid estimates of the \(Q\)s.
\item
  We hypothesize that the recommended thresholds to diagnose non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\), \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too stringent for iterative imputation applications. In an empirical study, where \(\widehat{R}\) was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet, scientific estimates were insensitive to continued iteration from \(t=5\) onward \citep{lace07}. We therefore suspect that \(\widehat{R}\) may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the bias the estimates. In contrast to this, it may occur that signs of non-convergence are under-estimated by \(\widehat{R}\), in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values. In practice, we do not expect this to cause problems for identifying non-convergence with \(\widehat{R}\).
\item
  We expect that high \(AC\) values are implausible in iterative imputation algorithms with typical convergence. That is, after only a few iterations, the randomness induced by the algorithm should effectively mitigate the risk of dependency within chains.
\item
  \textbf{{[}Something about thetas, that multivariate thetas are better at detecting non-convergence than univariate thetas. The non-convergence diagnostics applied on the chain means and chain variances will fail to identify signs of non-convergence when the multivariate Qs are affected.{]}}
\end{enumerate}

@@

In this study, sampling variance is not of interest. Therefore, a single complete dataset may serve as comparative truth in all simulation repetitions \citep{vink14}. The data-generating mechanism is a multivariate normal distribution, representing person-data on three predictor variables in a multiple linear regression problem \textbf{(from an unspecified social scientific field of study)}. Let the predictor space be defined as
\[
\begin{aligned}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{aligned}
\]
A finite population of \(N=1000\) is simulated using the \texttt{mvtnorm} package \citep{mvtnorm}. Subsequently, a fourth variable is constructed as outcome \(Y\). For each unit \(i = 1, 2,..., N\), let
\[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\]
where \(\epsilon \sim \mathcal{N}(0, 100)\). From the complete set, we obtain the true values of several quantities of scientific interest, \(Q\)s. \textbf{{[}MOVE this next part to the start of the Simulation Study section?{]} \(Q\)s are the targets that will be estimated in each simulation repetition, after first \emph{amputing} the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the \(Q\)s on the \(M\) completed datasets, and finally pooling the \(M\) estimates \(\hat{Q}\) to obtain a single \(\bar{Q}\) for each \(Q\). We try to predict the effect of non-convergence (i.e., the difference between \(Q\) and \(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and \(AC\), by applying these methods to several scalar summaries of the state space of the algorithm, \(\theta\)s.}

@@

\textbf{{[}Add to this section: 1) emphasize that the iterations in the plots are averages across repetitions, not within one run of MICE; 2) Add more info about figure legends and axes{]}}
